PanLex-RDF

### 2013-01-02 paper/2012/SWJ-PanLex/paper.tex
Title changed
made texttt environments small

\begin{itemize}
  \item \emph{Expressions} are lemmas. For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  They were extracted from the afore mentioned mono- or bilingual dictionaries or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language Varieties} within a language are further distinguished by arbitrary unique integer identifiers.
  For example, ``eng'' is the ISO 639-3 code for the English language.
  Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
  Their labels, when possible, are autonyms, written in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  Additionally these variety identifiers and labels do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} of expressions are distinguished by editors based on their
  interpretation of an information source.
%  Information sources can be dictionaries, thesauri, etc., and also personal knowledge. 
  The pair comprised of the editor and the information source is referred to as a meaning's \emph{approver}.
  Editors may add informal definitions to meanings in any language variety.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} relate expressions to meanings.
  So homonyms are represented by one expression being connected to different meanings.
  Denotations may be enriched with part-of-speech tags selected from a closed list based on the
  Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn.
  Furthermore, it is possible to generically tag denotations with metadata
  using key-value pairs.
  For instance, an English expression \emph{pig} when synonymous with
  \emph{police officer} could be annotated with \emph{pragmatics=vulgar}.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can be annotated with.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

@@

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
%The schema described above was adopted to create RDF triple structures.
%is a good starting point most of the relations were taken over introducing a PanLex RDF vocabulary.
The entities and relations of the schema described in the previous section
serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside
in the namespace \texttt{<http://panlex.org/plx/>}, abbreviated with \texttt{plx}.
An example of the resulting ontology is depicted in \autoref{tbl:vocabulary} and
summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern
\emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case
and the \{id\} is the primary key of the corresponding database table.

%As in the original PanLex database there are also \emph{expressions} used in the RDF dataset.
\begin{itemize}
\item Expressions are modeled as intances of \vocab{plx:Expression}.
Their textual representations and normalized versions thereof become the 
values of the properties \texttt{rdfs:label} and \texttt{plx:degradedText}, respectively.
Their corresponding language variety is stated
using \texttt{plx:languageVariety}.

\item For language and language varieties the classes
\texttt{plx:Language} and \texttt{plx:LanguageVariety} are introduced.
\emph{ISO 639-1} and \emph{ISO 639-3 codes} are of the classes
\texttt{plx:Iso639-1Code} and \texttt{plx:Iso639-3Code} respectively.

\item The RDF analog of the PanLex \emph{meaning} is the \texttt{plx:Meaning}.
Entities of this class may have an identifier assigned with the
\texttt{plx:identifier} property pointing to an \texttt{xsd:string} literal.
%This meaning identifier represents ids given to the considered meaning in other
%data sources and may at a later stage be used to build links to the meaning
%as represented on these sites.
Meanings may also have \emph{definitions}, entities of the
\texttt{plx:Definition} class, giving a textual representation
(\texttt{rdfs:label}) with more than just a lexeme in a certain
language variety (\texttt{plx:languageVariety}).

%These resources have an \texttt{rdfs:label} containing the textual representation of the considered expression.
%Additionally there is a normalized version of the label, reachable via the \texttt{plx:degradedText} property.
%Expression entities also have an assigned language variety (\texttt{plx:languageVariety}).
%It refers to the language variety the textual representation is written in.
\item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
These are entities of the \texttt{plx:Denotation} class pointing to meanings and expressions via the \texttt{plx:denotationMeaning} and \texttt{plx:denotationExpression} property respectively.
Denotations may also have a word class assigned to them.
This can be achieved with the denotation's \texttt{plx:wordClass} property pointing to a \texttt{plx:WordClass} entity.
\item All approvers share the \texttt{plx:Approver} class. The characteristics of an approver are described using mainly triples with literal objects.
These are for example \texttt{plx:registrationDate} pointing to an \texttt{xsd:date} literal,
\texttt{dc:title} to assign the title of a source,
\texttt{dc:creator} to give an \texttt{xsd:string} containing the author's name.
\end{itemize}

@@

Since the PanLex project compiled its database by extracting data from different sources, the licenses of these sources were also considered.
At present, we support the different license categories given in the database by creating resources of the \texttt{plx:License} class.
%with appropriate descriptions in their \texttt{rdfs:label}.
%In the RDF dataset these have an additional \texttt{rdfs:label} property giving the names or descriptions 
\begin{table}
\begin{scriptsize}
\begin{tabular}{ll}
Class & Properties \\
\toprule
\texttt{plx:Approver} & \texttt{plx:registrationDate} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{foaf:homepage} \\
                      & \texttt{plx:isbn} \\
                      & \texttt{dc:creator} \\
                      & \texttt{dc:title} \\
                      & \texttt{dc:publisher} \\
                      & \texttt{plx:yearOfPublication} \\
                      & \texttt{plx:quality} \\
                      & \texttt{plx:license} \\
\midrule
\texttt{plx:Language}
                      & \texttt{plx:iso639-3Code} \\
                      & \texttt{plx:iso639-1Code} \\
\midrule
\texttt{plx:LanguageVariety}
                      & \texttt{plx:languageVarietyOf} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Iso639-1Code} & \\
\midrule
\texttt{plx:Iso639-3Code} & \\
\midrule
\texttt{plx:Expression}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{plx:degradedText} \\
\midrule
\texttt{plx:Meaning}
                      & \texttt{plx:approver} \\
                      & \texttt{plx:meaningDefinition} \\
                      & \texttt{plx:identifier} \\
\midrule
\texttt{plx:Definition}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Denotation}
                      & \texttt{plx:denotationMeaning} \\
                      & \texttt{plx:denotationExpression} \\
                      & \texttt{plx:wordClass} \\
\midrule
\texttt{plx:WordClass}  
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:License}  & \texttt{rdfs:label} \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \texttt{rdf:type} and \texttt{rdfs:label} properties have been omitted for brevity.}
\label{tbl:vocabulary}
\end{table}

>>>>

\begin{itemize}
  \item \emph{Expressions} are lemmas. For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  They were extracted from the afore-mentioned mono- or bilingual dictionaries or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language varieties} within a language are further distinguished by arbitrary unique integer identifiers.
  For example, ``eng'' is the ISO 639-3 code for the English language.
  Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
  Their labels, when possible, are autonyms, written in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  Additionally these variety identifiers and labels do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} of expressions are distinguished by editors based on their
  interpretation of an information source.
%  Information sources can be dictionaries, thesauri, etc., and also personal knowledge. 
  The pair comprised of the editor and the information source is referred to as a meaning's \emph{approver}.
  Editors may add informal definitions to meanings in any language variety.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} relate expressions to meanings.
  So homonyms are represented by one expression being connected to different meanings.
  Denotations may be enriched with part-of-speech tags selected from a closed list based on the
  Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn.
  Furthermore, it is possible to generically tag denotations with metadata
  using key-value pairs.
  For instance, an English expression \emph{pig} when synonymous with
  \emph{police officer} could be annotated with \emph{pragmatics=vulgar}.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can be annotated with.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

@@

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
%The schema described above was adopted to create RDF triple structures.
%is a good starting point most of the relations were taken over introducing a PanLex RDF vocabulary.
The entities and relations of the schema described in the previous section
serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside
in the namespace \texttt{\small <http://panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{tbl:vocabulary} and
summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern
\emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case
and the \{id\} is the primary key of the corresponding database table.

%As in the original PanLex database there are also \emph{expressions} used in the RDF dataset.
\begin{itemize}
\item Expressions are modeled as intances of \vocab{plx:Expression}.
Their textual representations and normalized versions thereof become the 
values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
Their corresponding language variety is stated
using \texttt{\small plx:languageVariety}.

\item For language and language varieties the classes
\texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
\emph{ISO 639-1} and \emph{ISO 639-3 codes} are of the classes
\texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code} respectively.

\item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
Entities of this class may have an identifier assigned with the
\texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
%This meaning identifier represents ids given to the considered meaning in other
%data sources and may at a later stage be used to build links to the meaning
%as represented on these sites.
Meanings may also have \emph{definitions}, entities of the
\texttt{\small plx:Definition} class, giving a textual representation
(\texttt{\small rdfs:label}) with more than just a lexeme in a certain
language variety (\texttt{\small plx:languageVariety}).

%These resources have an \texttt{rdfs:label} containing the textual representation of the considered expression.
%Additionally there is a normalized version of the label, reachable via the \texttt{plx:degradedText} property.
%Expression entities also have an assigned language variety (\texttt{plx:languageVariety}).
%It refers to the language variety the textual representation is written in.
\item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the \texttt{\small plx:denotationMeaning} and \texttt{\small plx:denotationExpression} property respectively.
Denotations may also have a word class assigned to them.
This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
\item All approvers share the \texttt{\small plx:Approver} class. The characteristics of an approver are described using mainly triples with literal objects.
These are for example \texttt{\small plx:registrationDate} pointing to an \texttt{\small xsd:date} literal,
\texttt{\small dc:title} to assign the title of a source,
\texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name.
\end{itemize}

@@

Since the PanLex project compiled its database by extracting data from different sources, the licenses of these sources were also considered.
At present, we support the different license categories given in the database by creating resources of the \texttt{\small plx:License} class.
%with appropriate descriptions in their \texttt{rdfs:label}.
%In the RDF dataset these have an additional \texttt{rdfs:label} property giving the names or descriptions 
\begin{table}
\begin{scriptsize}
\begin{tabular}{ll}
Class & Properties \\
\toprule
\texttt{plx:Approver} & \texttt{plx:registrationDate} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{foaf:homepage} \\
                      & \texttt{plx:isbn} \\
                      & \texttt{dc:creator} \\
                      & \texttt{dc:title} \\
                      & \texttt{dc:publisher} \\
                      & \texttt{plx:yearOfPublication} \\
                      & \texttt{plx:quality} \\
                      & \texttt{plx:license} \\
\midrule
\texttt{plx:Language}
                      & \texttt{plx:iso639-3Code} \\
                      & \texttt{plx:iso639-1Code} \\
\midrule
\texttt{plx:LanguageVariety}
                      & \texttt{plx:languageVarietyOf} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Iso639-1Code} & \\
\midrule
\texttt{plx:Iso639-3Code} & \\
\midrule
\texttt{plx:Expression}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{plx:degradedText} \\
\midrule
\texttt{plx:Meaning}
                      & \texttt{plx:approver} \\
                      & \texttt{plx:meaningDefinition} \\
                      & \texttt{plx:identifier} \\
\midrule
\texttt{plx:Definition}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Denotation}
                      & \texttt{plx:denotationMeaning} \\
                      & \texttt{plx:denotationExpression} \\
                      & \texttt{plx:wordClass} \\
\midrule
\texttt{plx:WordClass}  
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:License}  & \texttt{rdfs:label} \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \texttt{\small rdf:type} and \texttt{\small rdfs:label} properties have been omitted for brevity.}
\label{tbl:vocabulary}
\end{table}
